Natural Selection in the 20th Century

### 2020-11-09 why-negative-selection.Rmd
Use earnings not household income.

---
title: "Natural Selection in the 20th Century"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "July 2020"
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
---

@@

@@

```{r, eval = FALSE}
# Open questions/TODO

* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1. 
* Reweight other regressions (e.g. the omnibus regression in the appendix)
* Look at geography, esp. in siblings regressions.
* Beauchamp 2016 used the HRS. He reckons it's representative. Is it? How do
  his effect sizes compare?
  - he also worries about the quality/quantity tradeoff, which might mean that
    people with fewer children have more grandchildren (than you'd expect). 
    As we have the two generations, we can test this. (Actually, there's a 
    positive correlation of about 0.1)

@@

```{r plot-means-over-time, fig.height = 5, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals. Green lines show a significant increase over time (p < 0.05). Red lines show a significant decrease."}

@@

n_regs <- length(score_names)
res_wt_combined %>% 
  mutate(
    Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                "Age/Qual/FLB", "None"),
    term     = pretty_names(term),
    term     = fct_reorder(term, estimate, order_abs(1))
  ) %>% 
  ggplot(aes(x = estimate, y = term, shape = Weights, fill = Weights, colour = p.value < 0.05/{{n_regs}})) +
  geom_point(size = 1.5, stroke = 0.5) +
  my_vline +
  pval_scale() +
  scale_fill_manual(values = c(
    "None"              = "grey15", 
    "Geographical"      = "dodgerblue4",
    "Age/Qualification" = "steelblue2",
    "Age/Qual/FLB"      = "darkseagreen4"
  )) +
  scale_shape_manual(values = c(
    "None" = "square filled", 
    "Geographical" = "circle filled",
    "Age/Qualification" = "circle filled", 
    "Age/Qual/FLB" = "circle filled"
  ))

@@

This explanation predicts that the strength of polygenic scores' effects on
fertility will correlate with the strength of their relationship with earnings.
To test this, we measure the correlation of each PGS with household income, 
as well as with education level, a phenotype which predicts earnings. We
then relate these measurements to effect sizes on number of children.
There is a clear negative relationship between a score's effect on
fertility, and its correlation with income or education (Figure \@ref(fig:plot-cors-income-educ)). If we control for age at first live birth,
the relationship changes sign and becomes positive.

```{r plot-cors-income-educ, fig.cap="Effect sizes on number of children, with and without controls for age at first live birth, by correlations with income and education"}

@@

cors_income <- cor(famhist[score_names], famhist$income_cat, use = "pairwise")
# TODO: use age_left_edu when this includes college
cors_educ <-   cor(famhist[score_names], famhist$edu_qual, use = "pairwise")

@@

dfr <- data.frame(
        Correlation  = rep(c("Income", "Education"), each = 33, 2),
        cor          = rep(c(cors_income[,1], cors_educ[,1]), 2),
        effect_size  = c(rep(effect_size, 2), rep(effect_size_flb, 2)),
        Controls     = rep(c("None", 
                        "Age at first live birth"), each = 66)
      )

>>>>

---
title: "Natural Selection in the 20th Century"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "July 2020"
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
---

@@

\normalem

@@

```{r TODO, eval = FALSE}
# Open questions/TODO

* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1. 
* Reweight other regressions (e.g. the omnibus regression in the appendix)?
* Look at geography, esp. in siblings regressions.
* We say we don't see change over time. But, we do see the effect of 
  controlling for AFLB changing over generations.
* Beauchamp 2016 used the HRS. He reckons it's representative. Is it? How do
  his effect sizes compare?
  - he also worries about the quality/quantity tradeoff, which might mean that
    people with fewer children have more grandchildren (than you'd expect). 
    As we have the two generations, we can test this. (Actually, there's a 
    positive correlation of about 0.1)
* Another point is that negative selection only among poor will increase 
    genetic inequality.
```

@@

```{r plot-means-over-time, fig.height = 5, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals. Green lines show a significant linear increase over time (p < 0.05/33). Red lines show a significant decrease."}

@@

n_regs <- length(score_names)
res_wt_combined %>% 
  mutate(
    Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                "Age/Qual/AFLB", "None"),
    term     = pretty_names(term),
    term     = fct_reorder(term, estimate, order_abs(1))
  ) %>% 
  ggplot(aes(x = estimate, y = term, shape = Weights, fill = Weights, colour = p.value < 0.05/{{n_regs}})) +
  geom_point(size = 1.5, stroke = 0.5) +
  my_vline +
  pval_scale() +
  scale_fill_manual(values = c(
    "None"              = "grey15", 
    "Geographical"      = "dodgerblue4",
    "Age/Qualification" = "steelblue2",
    "Age/Qual/AFLB"      = "darkseagreen4"
  )) +
  scale_shape_manual(values = c(
    "None" = "square filled", 
    "Geographical" = "circle filled",
    "Age/Qualification" = "circle filled", 
    "Age/Qual/AFLB" = "circle filled"
  ))

@@

This explanation predicts that the strength of polygenic scores' effects on
fertility will correlate with the strength of their relationship with earnings.
To test this, we calculate the correlation of each PGS with estimated
earnings from respondent's first job[^est-income], and with
 education level, a phenotype which predicts earnings. We
then relate these measurements to effect sizes on number of children.
There is a clear negative relationship between a score's effect on
fertility, and its correlation with earnings or education (Figure \@ref(fig:plot-cors-earnings-educ)). If we control for age at first live birth,
the relationship changes sign and becomes positive.

[^est-income]: Earnings are estimated from the 2007 Annual Survey of Hours and
Earnings, using the SOC 2000 job code.

```{r plot-cors-earnings-educ, fig.cap="Effect sizes on number of children, with and without controls for age at first live birth, by correlations with earnings and education. Each point represents one PGS."}

@@

cors_income <- cor(famhist[score_names], famhist$first_job_pay, use = "pairwise")
# TODO: use age_left_edu when this includes college
cors_educ <-   cor(famhist[score_names], famhist$edu_qual, use = "pairwise")

@@

dfr <- data.frame(
        Correlation  = rep(c("Earnings", "Education"), each = 33, 2),
        cor          = rep(c(cors_income[,1], cors_educ[,1]), 2),
        effect_size  = c(rep(effect_size, 2), rep(effect_size_flb, 2)),
        Controls     = rep(c("None", 
                        "Age at first live birth"), each = 66)
      )
